Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-07-10 paper.tex
Reformatted for sigplan.

\documentclass{llncs}

@@

\title{Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization}
\author{J. Ian Johnson and David Van Horn}
\institute{Northeastern University \\
           \email{\{ianj,dvanhorn\}@ccs.neu.edu}}

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. The
key to their method is to notice in languages without stack-capturing
features, the continuation is only modified a frame at a time. If the
state space without the stack is finite (and it can be made finite
with an approximation ala AAM), then the model falls directly into the
realm of pushdown systems. They thus recast the problem in terms of
pushdown systems. The whole of their machinery is then computing the
pushdown system on-the-fly, only considering states reachable from the
root (initial) state. To match nodes that push frames with later nodes
following the pop of that frame, they additionally compute an
$\epsilon$-closure graph, which keeps edges between nodes that have a
net-zero stack change path between them. We skip the detour to
pushdown systems altogether and show an operational understanding of
the resulting analysis.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization (see the last rule in figure
\ref{fig:summary-semantics}). Instead of following an entire path
through a call to return a value, we simply jump from the call to the
return with the result of the call (the second case of rule 4). We
will return to this analogy and what it means in the presence of
first-class control in section \ref{sec:sr}.

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \joinone{\mktab}{(\mpoint, \mstore')}{\mkont},
          \mmemo}$
\\
    & or \\
    & $\tpl{\mval_\mathit{result}, \mstore', \mkont, \mktab, \mmemo} \text{ if } \mval_\mathit{result} \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

Once we finitize the address space of the store, this computes a
characteristic finite state machine that has only reachable states of
the pushdown system the unbounded stack model embodies.  To turn this
semantics into PDCFA's algorithm for constructing a Dyck state graph,
we apply a widening operator ${\mathcal F}$ to make $\mstore$,
$\mkont$, and $\mmemo$ shared amongst all states.

\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

A $\System$ embodies the states seen and at which store, $S$, the
frontier set of states yet to analyze, $F$, the shared store for the
frontier, $\mstore$, the continuation table $\mktab$ and the memo
table $\mmemo$. All the states in the frontier are stepped with the
current store, after which the next store is the least upper bound of
all the resulting stores. The next frontier contains only states
resulting from stepping the previous frontier that we haven't seen at
this next store. Systematic techniques for a performant implementation
can be found in \citet{ianjohnson:oaam:2013}.

\paragraph{Example:} Let's consider a monovariant allocation strategy
(the address for each binding is the variable name itself) to run the
following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
We should expect that a pushdown analysis would predict this evaluates
to true, and there are no loops in the program. 0CFA claims there is a
loop from the second call of \texttt{id} to the first, and thusly
predicts this program evaluates to true or false. PDCFA claims there
is no loop, and depending on the implementation, that the result is
true or false (paper), or just true (implemented)\footnote{This is
  because the paper steps every seen state with the current store
  every iteration, but the implementation only steps states that need stepping.}. Let's take a look at the evaluation after binding
\texttt{id} ($\mstore_0 = [\texttt{id} \mapsto
\set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

\begin{enumerate}
\item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
    \mstore_0[\texttt{x} \mapsto \set{0}]$ and
    (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
         $\mkont_1$ = \texttt{(letk y (let ([z (id 1)]) ($\le$ y z)) $[\texttt{id}\mapsto{\texttt{id}}]$)})
    $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
\item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
\item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
\item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \texttt{(letk z ($\le$ y z) $[\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]$)} (call this $\mkont_2$).}
\item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
      (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
\item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
\item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
\end{enumerate}

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of a nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

@@

\begin{align*}
  \bind(\mstore,\mframe,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{ll}
            (\mstore, \snglm{\maddr}{\mval}) & \text{if } \mvar \text{ never escapes} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) & \text{otherwise } \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{ll}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))}
                   & \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} & \text{otherwise}
           \end{array}\right.
\end{align*}

@@

There is contention among programming language researchers whether
\rackett{call/cc} should be a language primitive, since it captures
the entire stack, leading to space leaks
\citep{ianjohnson:kiselyov:against-callcc}. Alternative control
operators have been proposed that delimit how much of the stack to
capture, such as Felleisen's $\%$ (read ``prompt'') and capture
operator ${\mathcal F}$ (read ``control'')
\citep{ianjohnson:felleisen:control:1988}, or Danvy and Filinski's
\texttt{reset} (equivalent to $\%$) and \texttt{shift}
\citep{ianjohnson:danvy:filinski:delim:1990}. However, the stacks
captured by these operators always extend the stack when invoked,
rather than replace it like those captured with
\rackett{call/cc}. Continuations that have this extension behavior are
called ``composable continuations.'' Stack replacement is simple to
model in a regular analysis using the AAM approach, and Vardoulakis
and Shivers showed it can be done in a pushdown approach (although it
breaks the pushdown model). Stack extension, however, poses a new
challenge for pushdown analysis, since one application of a composable
continuation means pushing an unbounded amount of frames onto the
stack. Vardoulakis' and Shivers' approach does not immediately apply
in this situation, since their technique drops all knowledge of the
stack at a continuation's invocation site; extension, however, must
preserve it.

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}}$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\mkont},\kmt,\mmkont}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\mkont', \mkont}, \mmkont}$
    &
    $\tpl{\mval, \mstore, \mkont', \mkapp{\mkont}{\mmkont}}$
  \end{tabular}  
  \caption{Machine semantics for shift/reset}
  \label{fig:shift-reset}
\end{figure}

@@

\begin{align*}
  \mctx \in \Context &::= ((\mexpr,\menv),\mstore) \mid (\vcomp{\mkont},\mval,\mstore) \\
  \mkont \in \Kont &= \kmt \mid \krt{\mctx} \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \mmkont \in \MKont &::= \kmt \mid \kprompt{\mctx} \\
  \mktab \in \KTab &= \Context \to (\wp(\Kont \times \MKont) \cup \Kont) \\
  \mmemo \in \Memo &= \Context \to \wp(\Value)
\end{align*}

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mpoint, \mstore, \kmt, \kprompt{\mctx}, \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}, \mmemo}$
    \\
    where & $\mpoint = (\mexpr, \menv)$, $\mctx = (\mpoint, \mstore)$.
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \kprompt{\mctx}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}, \mktab, \joinone{\mmemo}{\mctx}{\mval}}$
    if $(\mkont,\mmkont) \in \mktab(\mctx)$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont}},\kmt,\mmkont,\mktab,\mmemo}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont', \kprompt{\mctx}, \mktab',\mmemo}$
    \\
    where & $\mctx = (\mkont, \mval, \mstore')$ \\
          & $\mktab' = \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}$
    \\
% Memoized continuation call
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval', \mstore, \mkont, \mmkont, \mktab,\mmemo}$ if $\mval' \in \mmemo(\mkont',\mval,\mstore)$
  \end{tabular}  
  \caption{Faulty table-based semantics for shift/reset}
  \label{fig:shift-reset-table0}
\end{figure}

@@

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. In particular, we would want to have an
escape analysis to find first-class continuations that don't need to
be heap-allocated, and single target abort operations that can be
turned into long jumps after a computed stack unraveling. On top of
optimizations, there are security analyses. Greater control of the
stack gives us the ability to drill deep into context-sensitive
security properties and make precise predictions such as, ``is it
possible to send image data on the network outside the context of the
user's permission?'' This way a user can use an image taking/sharing
application without the application surreptitiously taking and sending
images the user doesn't know about.

\bibliographystyle{chicago}
\bibliography{bibliography}

>>>>

\documentclass{sigplanconf}

@@

\title{Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization}
\authorinfo{J. Ian Johnson and David Van Horn}{Northeastern University}{\{ianj,dvanhorn\}@ccs.neu.edu}

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. The
key to their method is to notice in languages without stack-capturing
features, the continuation is only modified a frame at a time. If the
state space without the stack is finite (and it can be made finite
with an approximation ala AAM), then the model falls directly into the
realm of pushdown systems. They thus recast the problem in terms of
pushdown systems. The whole of their machinery is then computing the
pushdown system on-the-fly, only considering states reachable from the
root (initial) state. They call the main data-structure for this a
Dyck state graph. To match nodes that push frames with later nodes
following the pop of that frame, they additionally compute an
$\epsilon$-closure graph, which keeps edges between nodes that have a
net-zero stack change path between them. We skip the detour to
pushdown systems altogether and show an operational understanding of
the resulting analysis.

@@

\paragraph{The role of summaries:} notice the additional $\Memo$
component. A ``summary edge'' in CFA2 or equivalently an
$\epsilon$-edge in PDCFA is an edge from the source of a push edge to
the target of the matching pop edge. ``Matching'' here means there is
a path through machine reductions that don't change the stack, or
through summary edges. In our model, we only ``push'' when we call a
function, and we only ``pop'' when we return with a value. There is an
analogy to something more operational: summary edges embody
memoization (see the last rule in figure
\ref{fig:summary-semantics}). Instead of following an entire path
through a call to return a value, we simply jump from the call to the
return with the result of the call (the second case of rule 4).

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% variable lookup
    $\tpl{(\svar\mvar, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{\mval,\mstore,\mkont, \mktab, \mmemo}$ if $\mval \in \mstore(\menv(\mvar))$
    \\
% application
    $\tpl{(\sapp{\mexpri0}{\mexpri1}, \menv), \mstore, \mkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpri0, \menv), \mstore, \kar{\mexpri1,\menv,\mkont}, \mktab, \mmemo}$
    \\
% argument evaluation
    $\tpl{\mval, \mstore, \kar{\mexpr,\menv,\mkont}, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kfn{\mval, \mkont}, \mktab, \mmemo}$
    \\
% function call
    $\tpl{\mval,\mstore,\kfn{\vclo{\slam{\mvar}{\mexpr}}{\menv}, \mkont},\mktab,\mmemo}$
    & % actually do call,
    $\tpl{\mpoint,
          \mstore',
          \krt{\mpoint, \mstore'},
          \mktab',
          \mmemo}$
\\
    & or \\
    & $\tpl{\mval_\mathit{result},
            \mstore',
            \mkont,
            \mktab',
            \mmemo}$
    \\ & \quad if $\mval_\mathit{result} \in \mmemo(\mpoint,\mstore')$
    \\ % or, lookup in table
%    & $\tpl{\mval, \mstore', \mkont,\mktab,\mmemo}$ if $\mval \in \mmemo(\mpoint, \mstore')$ \\    
    where & $\mpoint = (\mexpr, \extm{\menv}{\mvar}{\maddr})$ \\
          & $\mstore' = \joinone{\mstore}{\maddr}{\mval}$ \\
          & $\mktab' = \joinone{\mktab}{(\mpoint, \mstore')}{\mkont}$
    \\
% return
    $\tpl{\mval, \mstore, \krt{\mpoint,\mstore'}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont, \mktab, \joinone{\mmemo}{(\mpoint, \mstore')}{\mval}}$
    \\ & \quad if $\mkont \in \mktab(\mpoint, \mstore')$
  \end{tabular}
  \caption{The summarizing tabular stack machine}
  \label{fig:summary-semantics}
\end{figure}

To turn this semantics into PDCFA's algorithm for constructing a Dyck
state graph, given the appropriate $\alloc$, we apply a widening
operator to make $\mstore$, $\mkont$, and $\mmemo$ shared amongst all
states (meaning states then become represented without these
compontents), in what is then called a $\System$.  Since these
components are shared, they are always the least upper bound of any
respective component that the semantics produces (the intermediate set
$I$) in order to stay sound. We use a meta-function $\wn$, ``wide to
narrow'', to shift between the two representations of states.

\begin{align*}
  \mastate \in \sa{State} &= (\Expr \times \Env) \times \Kont \\
  \System &= \wp(\sa{State} \times \Store) \times \wp(\sa{State}) \\
          &\phantom{=} \times \Store \times \KTab \times \Memo \\
  {\mathcal F} &: \System \to \System \\
  {\mathcal F}(S, F, \mstore, \mktab, \mmemo) &= (S \cup S', F', \mstore', \mktab', \mmemo') \\
  \text{where } I &= \setbuild{\mstate'}{\mastate \in F, \wn(\mastate,\mstore,\mktab,\mmemo) \stepto \mstate'} \\
                \mstore' &= \bigsqcup\setbuild{\mstore'}{\wn(\_, \mstore',\_,\_) \in I} \\
                \mktab' &=  \bigsqcup\setbuild{\mktab'}{\wn(\_, \_, \mktab', \_) \in I} \\
                \mmemo' &=  \bigsqcup\setbuild{\mmemo'}{\wn(\_, \_, \_, \mmemo') \in I} \\
                S' &= \setbuild{(\mastate',\mstore')}{\wn(\mastate', \_, \_, \_) \in I} \\
                F' &= \setbuild{\mastate'}{(\mastate',\_) \in S' \setminus S} \\
                \wn(\tpl{(\mexpr,\menv),\mkont},\mstore,\mktab,\mmemo)
                   &= \tpl{(\mexpr,\menv),\mstore,\mkont,\mktab,\mmemo}
\end{align*}

A $\System$ embodies the states seen and at which store, $S$, the
frontier set of states yet to analyze, $F$, the shared store for the
frontier, $\mstore$, the continuation table $\mktab$ and the memo
table $\mmemo$. All the states in the frontier are stepped with the
current store, after which the next store is the least upper bound of
all the resulting stores. The next frontier contains only states
resulting from stepping the previous frontier that we haven't seen at
this next store. Systematic techniques for a performant implementation
can be found in \citet{DBLP:journals/corr/abs-1211-3722}.

\paragraph{Example:} Let's consider a monovariant allocation strategy
(the address for each binding is the variable name itself) to run the
following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
Suppose we extend our semantics to allow numbers, numeric primitives
and \texttt{let}. The continuation frame for a \texttt{let} contains
the identifier to bind to the resulting value, along with the body of
the \texttt{let} with its environment. Call the constructor of this
frame $\mathbf{lt}$.

We should expect that a pushdown analysis would predict this evaluates
to true, and there are no loops in the program. 0CFA ~\citep{ianjohnson:Shivers:1991:CFA} claims there is a
loop from the second call of \texttt{id} to the first, and thusly
predicts this program evaluates to true or false. PDCFA claims there
is no loop, and depending on the implementation, that the result is
true or false (paper), or just true (implemented)\footnote{This is
  because the paper steps every seen state with the current store
  every iteration, but the implementation only steps states that need stepping.}. Let's take a look at the evaluation after binding
\texttt{id} ($\mstore_0 = [\texttt{id} \mapsto
\set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

\begin{enumerate}
\item{\texttt{(id 0)} steps to \texttt{x} at $\krt{((\texttt{x}, [\texttt{x}\mapsto\texttt{x}]), \mstore_0)}$ with $\mstore_1 =
    \mstore_0[\texttt{x} \mapsto \set{0}]$ and
    (let $\mctx_1 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_1)$,
         $\mkont_1 = \klet{\texttt{y}, \texttt{(let ([z (id 1)]) ($\le$ y z))}, [\texttt{id}\mapsto{\texttt{id}}]}$)
    $\mktab_1 = [\mctx_0 \mapsto \set{\mkont_0}]$}
\item{\texttt{0} at $\krt{\mctx_1}$ steps to $0$ at $\mkont_1$ and $\mmemo_1 = [\mctx_1 \mapsto \set{0}]$}
\item{\texttt{0} at $\mkont_1$ steps to \texttt{(let ([z (id 1)]) ($\le$ y z))} and $\mstore_2 = \mstore_1[\texttt{y} \mapsto \set{0}]$}
\item{\texttt{(let ([z (id 1)]) ($\le$ y z))} steps to \texttt{(id 1)} at \\
      $\klet{\texttt{z}, \texttt{($\le$ y z)}, [\texttt{id} \mapsto \texttt{id}, \texttt{y} \mapsto \texttt{y}]}$
      (call this $\mkont_2$).}
\item{\texttt{(id 1)} steps to $x$ with $\mstore_3 = \mstore_2[\texttt{x} \mapsto \set{0,1}]$ and
      (let $\mctx_3 = ((\texttt{x},[\texttt{x}\mapsto\texttt{x}]),\mstore_3)$) $\mktab_3 =\mktab_1[\mctx_3 \mapsto \set{\mkont_2}]$.}
\item{\texttt{0} or \texttt{1} at $\krt{\mctx_3}$ steps to \texttt{0} or \texttt{1} at $\mkont_2$ and $\mmemo_3 = \mmemo_1[\mctx_3 \mapsto \set{0,1}]$.}
\item{\texttt{z} gets bound to $\set{0,1}$, and \texttt{($\le$ y z)} evaluates to true.}
\end{enumerate}

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings in an additional $\mframe \in \Store$}
\item{strong updates on stack frames for resolved non-determinism}
\end{enumerate}
The first of these is an addition to the stack-less context. There is
a conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references (labeled with a distinguishing
$\mlab$ from an arbitrary space of labels) as able to use the stack frame or not. Their criteria for a
binding never escaping is that it is never referenced in a function
that is not its binder. This can be extended in a language with more
linguistic features; see Kranz's thesis about register-allocatable
bindings in the Orbit Scheme compiler
\citep{ianjohnson:kranz:thesis:1988}. A stackable reference is one
that appears in the body of the binding function, by which we mean not
within the body of a nested function. They use the information that a
binding never escapes to not bother allocating it in the heap. This
has the advantage of not changing the heap, and thus leads to less
propagation. The addition of these stack frames makes the analysis
exponential in theory, though in practice they have been observed to
decrease running time in most cases.

@@

\begin{align*}
  \bind(\mstore,\mframe,\maddr,\mvar,\mval) &=
   \left\{\begin{array}{l}
            (\mstore, \snglm{\maddr}{\mval}) \text{ if } \mvar \text{ never escapes} \\
            (\joinone{\mstore}{\maddr}{\mval}, \snglm{\maddr}{\mval}) \text{ otherwise} \\
          \end{array}\right. \\
  \lookup(\mstore,\mframe,\menv,\mvar,\mlab) &=
    \left\{\begin{array}{l}
          \setbuild{(\extm{\mframe}{\menv(\mvar)}{\set{\mval}}, \mval)}
                   {\mval \in \mframe(\menv(\mvar))}
                   \\ \qquad \text{if } \mlab \text{ non-escaping} \\
          \setbuild{(\mframe,\mval)}{\mval \in \mstore(\menv(\mvar))} \text{ otherwise}
           \end{array}\right.
\end{align*}

@@

There is contention among programming language researchers whether
\rackett{call/cc} should be a language primitive, since it captures
the entire stack, leading to space leaks
\citep{ianjohnson:kiselyov:against-callcc}. Alternative control
operators have been proposed that delimit how much of the stack to
capture, such as $\%$ (read ``prompt'') and capture operator
${\mathcal F}$ (read ``control'')
\citep{ianjohnson:felleisen:control:1988}, or \texttt{reset}
(equivalent to $\%$) and \texttt{shift}
\citep{ianjohnson:danvy:filinski:delim:1990}. However, the stacks
captured by these operators always extend the stack when invoked,
rather than replace it like those captured with
\rackett{call/cc}. Continuations that have this extension behavior are
called ``composable continuations.'' Stack replacement is easily
modeled in a regular analysis using the AAM approach, and Vardoulakis
and Shivers showed it can be done in a pushdown approach (although it
breaks the pushdown model). Stack extension, however, poses a new
challenge for pushdown analysis, since one application of a composable
continuation means pushing an unbounded amount of frames onto the
stack. Vardoulakis' and Shivers' approach does not immediately apply
in this situation, since their technique drops all knowledge of the
stack at a continuation's invocation site; extension, however, must
preserve it.

@@

\begin{figure}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \menv), \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \mkapp{\mkont}{\mmkont}}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}}$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \mstore',\kmt,\mmkont}$
    \\ & where $\mstore' = \joinone{\mstore}{\maddr}{\mkont}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\mkont', \mkont}, \mmkont}$
    &
    $\tpl{\mval, \mstore, \mkont', \mkapp{\mkont}{\mmkont}}$
  \end{tabular}  
  \caption{Machine semantics for shift/reset}
  \label{fig:shift-reset}
\end{figure}

@@

\begin{align*}
  \mctx \in \Context &::= ((\mexpr,\menv),\mstore) \mid (\vcomp{\mkont},\mval,\mstore) \\
  \mkont \in \Kont &= \kmt \mid \krt{\mctx} \mid \kar{\mexpr,\menv,\mkont} \mid \kfn{\mval,\mkont} \\
  \mmkont \in \MKont &::= \kmt \mid \kprompt{\mctx} \\
  \mathit{KTabRange} &= \wp(\Kont \times \MKont) \cup \wp(\Kont) \\
  \mktab \in \KTab &= \Context \to \mathit{KTabRange} \\
  \mmemo \in \Memo &= \Context \to \wp(\Value)
\end{align*}

@@

\begin{figure*}
  \centering
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
% Reset
    $\tpl{(\sreset{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mpoint, \mstore, \kmt, \kprompt{\mctx}, \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}, \mmemo}$
    \\
    where & $\mpoint = (\mexpr, \menv)$, $\mctx = (\mpoint, \mstore)$.
    \\
% Pop prompt
    $\tpl{\mval, \mstore, \kmt, \kprompt{\mctx}, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, {\mkont}, {\mmkont}, \mktab, \joinone{\mmemo}{\mctx}{\mval}}$
    if $(\mkont,\mmkont) \in \mktab(\mctx)$
    \\
% Shift
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont}},\kmt,\mmkont,\mktab,\mmemo}$
    \\
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval, \mstore, \mkont', \kprompt{\mctx}, \mktab',\mmemo}$
    \\
    where & $\mctx = (\mkont', \mval, \mstore)$ \\
          & $\mktab' = \joinone{\mktab}{\mctx}{(\mkont,\mmkont)}$
    \\
% Memoized continuation call
% continuation call
    $\tpl{\mval,\mstore,\kfn{\vcomp{\mkont'}, \mkont}, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{\mval', \mstore, \mkont, \mmkont, \mktab,\mmemo}$ if $\mval' \in \mmemo(\mkont',\mval,\mstore)$
  \end{tabular}  
  \caption{Faulty table-based semantics for shift/reset}
  \label{fig:shift-reset-table0}
\end{figure*}

@@

There is also the question of clients of these analyses. Most
obviously we would want to know where we can implement first-class
control more efficiently. In particular, we would want to have an
escape analysis to find first-class continuations that don't need to
be heap-allocated, and single target abort operations that can be
turned into long jumps after a computed stack unraveling. On top of
optimizations, there are security analyses. Greater control of the
stack gives us the ability to drill deep into context-sensitive
security properties and make precise predictions.

\bibliographystyle{plainnat}
\bibliography{bibliography}
